---
title: "Homework 1"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "February 2, 2016"
output: pdf_document
geometry: margin=0.5in
---
```{r, echo=FALSE}
library(knitr)
set.seed(20275)
```

## 1.1

Name several entities, attributes, activities, events, and state variables for the following systems.

### (a) A cafeteria

#### Entities

* Serving Line
* Food Servers
* Tables

#### Attributes

* Number of Food Servers
* Number of seats per table
* Rate of serving for Food Servers
* Time range for eating the meal

#### Activities

* Waiting in line
* Being served by a Food Server
* Waiting for a table to eat
* Eating at a table

#### Events

* Arrival of new person in serving line to be served
* Person leaving serving line
* Person waiting for seat at table to eat
* Person finishing eating and leaving table

#### State Variables

* Number of people eating at tables
* Number of people waiting in line to be served

### (b) A grocery store

#### Entities

* Checkout lanes

#### Attributes

* Max number of items allowed in checkout lane
* Rate of checkout for cashier

#### Activities

* Customer shopping in the grocery store
* Customer checking out (paying for goods)

#### Events

* Arrival of customer at grocery store
* Arrival of customer at checkout lane
* Customer completing checkout
* Customer departing store without purchasing anything

#### State Variables

* Number of customers in grocery store
* Number of customers in checkout lane lines


### (c) A laundromat

#### Entities

* Washing machines
* Drying machines

#### Attributes

* Washing machine run time
* Drying machine run time
* Ratio of washing machine to drying machine capacity

#### Activities

* Washing clothes
* Drying clothes
* Loading washing machine
* Transfering from washing to drying machine
* Unloading from drying machine

#### Events

* Washing maching cycle starts
* Washing machine cycle stops
* Dryer cycle starts
* Dryer cycle stops

#### State Variables

* Number of busy washing machines
* Number of busy dryers

### (d) A fast-food restaurant

#### Entities

* Cashiers
* Back-cooks (i.e. burger flippers)
* Fryers

#### Attributes

* Burgers per burger flipper
* Orders of frys per Fryer
* Cashier busy or not

#### Activities

* Cooking a burger
* Making french frys
* Cashier taking order, accepting payment

#### Events

* Order in
* Order ready for pickup
* French fries done cooking

#### State Variables

* Number of orders pending
* Number of burgers being cooked
* Orders of french frys cooked/ready for serving
* Number of bugers being ready for serving

### (e) A hospital emergency room

#### Entities

* Doctors
* Beds
* Patients
* Admitting staff

#### Attributes

* Patients per Doctor


#### Activities

* Patient admitted
* Doctor take care of patient
* Patient discharged

#### Events

* Patient arrives
* Patient admitted
* Doctor discharges patient

#### State Variables

* Beds empty
* Patients awaiting addmission
* Paitents awaiting discharge


### (f) A taxicab company with 10 taxis

#### Entities

* Taxis
* Dispatcher
* Customers

#### Attributes

* Taxi has customer
* Taxi enroute to customer
* Customer waiting for taxi

#### Activities

* Enroute to customer
* Transporting customer

#### Events

* Picking up custommer
* Dropping off customer

#### State Variables

* Taxis with customers
* Customers waiting for available taxi

### (g) An automobile assembly line

#### Entities

* Parts
* Assembly machines
* Workers

#### Attributes

* Parts inventory
* Assembly machine rate of production
* Worker rate of production

#### Activities

* Machine assembling car 
* Worker assembling car
* Staging parts for use by Machine or Worker

#### Events

* Car assembly started
* Car assembly completed
* Parts depleted
* Car assembly by machine X completed
* Car assembly by worker Y completed

#### State Variables

* Cars on assembly line
* Parts inventory level
* Workers out sick/vacation
* Machines broken down

## 2.1 

Consider the following continuously operating job shop. Interarrival times of jobs are distributed as follows:

Time Between Arrivals (hours) | Probability
------------------------------|------------
0                             |  0.23
1                             |  0.37
2                             |  0.28
3                             |  0.12

Processing times for jobs are normally distributed, with mean 50 minutes, and standard deviation 8 minutes. Construct
a simulation table and perform a simulation for 10 new customers. Assume that, when the simulation starts, there is one
job being processed (scheduled to be completed in 25 minutes) and there is one job with a 50-minute processing time in
the queue.


```{r}
# Create a data frame of the pre-existing jobs
existingJobs <- data.frame(customer=c(-2, -1), 
                           iaHrs=c(0,0), 
                           iaMins=c(0,0), 
                           arrivalMins=c(0,0),
                           svcTimeMins=c(25, 50),
                           timeSvcBegin=c(0, 25),
                           queueWaitMins=c(0,25),
                           timeSvcEnd=c(25, 75),
                           timeInSystem=c(25,75))
# Create a data frame of the new customers and their jobs
newJobs <- data.frame(customer=seq(1, 10), 
                      iaHrs=c(0, sample(seq(0, 3), 
                                              size=9, 
                                              prob=c(.23, .37, .28, .12), 
                                              replace=TRUE)), 
                      iaMins=rep(NA, 10), 
                      arrivalMins=rep(0, 10),
                      svcTimeMins=rnorm(10, mean=50, sd=8),
                      timeSvcBegin=rep(0, 10),
                      queueWaitMins=rep(0, 10),
                      timeSvcEnd=rep(0, 10),
                      timeInSystem=rep(0, 10))
# Convert from interarrival hours to minutes and 
# determine overall arrival times
newJobs$iaMins <- newJobs$iaHrs * 60
newJobs$arrivalMins <- cumsum(newJobs$iaMins) 
# Join the existing and new jobs into one table
simTable <- rbind(existingJobs, newJobs)
# Loop over the rows the compute the various activity and clock times
for(i in seq(3, nrow(simTable)))
{
  simTable[i,]$timeSvcBegin <- max(simTable[i,]$arrivalMins, simTable[i-1,]$timeSvcEnd)
  simTable[i,]$queueWaitMins <- simTable[i,]$timeSvcBegin - simTable[i,]$arrivalMins
  simTable[i,]$timeSvcEnd <- simTable[i,]$timeSvcBegin + simTable[i,]$svcTimeMins
  simTable[i,]$timeInSystem <- simTable[i,]$timeSvcEnd - simTable[i,]$arrivalMins
}
# Show the table
kable(simTable)
```

#### (a) What was the average time in the queue for the 10 new jobs?

The average time in the queue for the 10 new jobs is computed below:

```{r}
mean(simTable[seq(3, 12),]$queueWaitMins)
```

#### (b) What was the average processing time of the 10 new jobs?

The average processing time is computed below:

```{r}
mean(newJobs$svcTimeMins)
```


#### (c) What was the maximum time in the system for the 10 new jobs?

The maximum time in the system for the 10 new jobs is computed below:

```{r}
max(simTable[seq(3, 12),]$timeInSystem)
```

## 2.2

A baker is trying to figure out how many dozens of bagels to bake each day. The probability distribution of the number of bagel customers is as follows:

Customer/Day | 8    | 10   | 12   | 14 
-------------|------|------|------|----
Probability  | 0.35 | 0.30 | 0.25 | 0.10

Customers order 1,2,3 or 4 dozen bagels according to the following probability distribution:

Dozen Ordered/Customer | 1   | 2   | 3   | 4   
-----------------------|-----|-----|-----|-----
Probability            | 0.4 | 0.3 | 0.2 | 0.1

Bagels sell for $8.40 per dozen. They cost $5.80 per dozen to make. All bagels not sold at the end of the day
are sold at half priec to a local grocery store. Based on 5 days of simulation, how many dozen (to the nearest 5 dozen)
bagels should be baked each day?

```{r}
simDays <- 5
revPerDoz <- 8.40
costPerDoz <- 5.80
bagelsMade <- 15
simTable <- data.frame(day=seq(1, simDays),
                       customers=sample(c(8,10,12,14), 
                                        size=simDays, 
                                        prob=c(0.35, 0.30, 0.25, 0.10),
                                        replace=TRUE),
                       dozenOrdered=rep(NA, simDays),
                       revenue=rep(NA, simDays),
                       lostProfit=rep(NA, simDays),
                       salvage=rep(NA, simDays),
                       dailyCost=rep(NA, simDays),
                       dailyProfit=rep(NA, simDays))

for(i in seq(1, nrow(simTable)))
{
  bagelsOrdered <- sample(c(1,2,3,4),
                          size=simTable[i,]$customers,
                          prob=c(0.4, 0.3, 0.2, 0.1),
                          replace=TRUE)
  
  simTable[i,]$dozenOrdered <- sum(bagelsOrdered)
  
  simTable[i,]$revenue <- min(simTable[i,]$dozenOrdered, bagelsMade) * revPerDoz
  simTable[i,]$lostProfit <- max(simTable[i,]$dozenOrdered - bagelsMade, 0) * (revPerDoz - costPerDoz)
  simTable[i,]$salvage <- max(bagelsMade - simTable[i,]$dozenOrdered, 0) * (revPerDoz / 2)
  simTable[i,]$dailyCost <- bagelsMade * costPerDoz
  simTable[i,]$dailyProfit <- simTable[i,]$revenue + simTable[i,]$salvage - simTable[i,]$dailyCost
  
}

kable(simTable)

```

5 day profit is `r sum(simTable$dailyProfit)` at `r bagelsMade` dozen bagels made per day.