---
title: "Homework 1"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "February 2, 2016"
output: pdf_document
---
```{r, echo=FALSE}
set.seed(20275)
```

## 1.1

Name several entities, attributes, activities, events, and state variables for the following systems.

### (a) A cafeteria

#### Entities

* Serving Line
* Food Servers
* Tables

#### Attributes

* Number of Food Servers
* Number of seats per table
* Rate of serving for Food Servers
* Time range for eating the meal

#### Activities

* Waiting in line
* Being served by a Food Server
* Waiting for a table to eat
* Eating at a table

#### Events

* Arrival of new person in serving line to be served
* Person leaving serving line
* Person waiting for seat at table to eat
* Person finishing eating and leaving table

#### State Variables

* Number of people eating at tables
* Number of people waiting in line to be served

### (b) A grocery store

#### Entities

* Checkout lanes

#### Attributes

* Max number of items allowed in checkout lane
* Rate of checkout for cashier

#### Activities

* Customer shopping in the grocery store
* Customer checking out (paying for goods)

#### Events

* Arrival of customer at grocery store
* Arrival of customer at checkout lane
* Customer completing checkout
* Customer departing store without purchasing anything

#### State Variables

* Number of customers in grocery store
* Number of customers in checkout lane lines


### (c) A laundromat

#### Entities

* Washing machines
* Drying machines

#### Attributes

* Washing machine run time
* Drying machine run time
* Ratio of washing machine to drying machine capacity

#### Activities

* Washing clothes
* Drying clothes
* Loading washing machine
* Transfering from washing to drying machine
* Unloading from drying machine

#### Events

* Washing maching cycle starts
* Washing machine cycle stops
* Dryer cycle starts
* Dryer cycle stops

#### State Variables

* Number of busy washing machines
* Number of busy dryers

### (d) A fast-food restaurant

#### Entities

* Cashiers
* Back-cooks (i.e. burger flippers)
* Fryers

#### Attributes

* Burgers per burger flipper
* Orders of frys per Fryer
* Cashier busy or not

#### Activities

* Cooking a burger
* Making french frys
* Cashier taking order, accepting payment

#### Events

* Order in
* Order ready for pickup
* French fries done cooking

#### State Variables

* Number of orders pending
* Number of burgers being cooked
* Orders of french frys cooked/ready for serving
* Number of bugers being ready for serving

### (e) A hospital emergency room

#### Entities

* Doctors
* Beds
* Patients
* Admitting staff

#### Attributes

* Patients per Doctor


#### Activities

* Patient admitted
* Doctor take care of patient
* Patient discharged

#### Events

* Patient arrives
* Patient admitted
* Doctor discharges patient

#### State Variables

* Beds empty
* Patients awaiting addmission
* Paitents awaiting discharge


### (f) A taxicab company with 10 taxis

#### Entities

* Taxis
* Dispatcher
* Customers

#### Attributes

* Taxi has customer
* Taxi enroute to customer
* Customer waiting for taxi

#### Activities

* Enroute to customer
* Transporting customer

#### Events

* Picking up custommer
* Dropping off customer

#### State Variables

* Taxis with customers
* Customers waiting for available taxi

### (g) An automobile assembly line

#### Entities

* Parts
* Assembly machines
* Workers

#### Attributes

* Parts inventory
* Assembly machine rate of production
* Worker rate of production

#### Activities

* Machine assembling car 
* Worker assembling car
* Staging parts for use by Machine or Worker

#### Events

* Car assembly started
* Car assembly completed
* Parts depleted
* Car assembly by machine X completed
* Car assembly by worker Y completed

#### State Variables

* Cars on assembly line
* Parts inventory level
* Workers out sick/vacation
* Machines broken down

2.1 Consider the following continuously operating job shop. Interarrival times of jobs are distributed as follows:

Time Between Arrivals (hours) | Probability
------------------------------|------------
0                             |  0.23
1                             |  0.37
2                             |  0.28
3                             |  0.12

Processing times for jobs are normally distributed, with mean 50 minutes, and standard deviation 8 minutes. Construct
a simulation table and perform a simulation for 10 new customers. Assume that, when the simulation starts, there is one
job being processed (scheduled to be completed in 25 minutes) and there is one job with a 50-minute processing time in
the queue.


```{r}
existingJobs <- data.frame(customer=c(-2, -1), 
                           interarrivalHrs=c(0,0), 
                           interarrivalMins=c(0,0), 
                           arrivalMins=c(0,0),
                           jobProcessingMins=c(25, 50))

newJobs <- data.frame(customer=seq(1, 10), 
                       interarrivalHrs=sample(seq(0, 3), 
                                              size=10, 
                                              prob=c(.23, .37, .28, .12), 
                                              replace=TRUE), 
                       interarrivalMins=rep(NA, 10), 
                       arrivalMins=rep(0, 10),
                       jobProcessingMins=rnorm(10, mean=50, sd=8))

newJobs$interarrivalMins <- newJobs$interarrivalHrs * 60
newJobs$arrivalMins <- cumsum(newJobs$interarrivalMins) 

simTable <- rbind(existingJobs, newJobs)
simTable

```

#### (a) What was the average time in the queue for the 10 new jobs?


#### (b) What was the average processing time of the 10 new jobs?

The average processing time is computed below:

```{r}
mean(newJobs$jobProcessingMins)
```


#### (c) What was the maximum time in the system for the 10 new jobs?