---
title: "Homework 1"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "February 2, 2016"
output: pdf_document
classoption: landscape
geometry: margin=0.5in
---
```{r, echo=FALSE}
library(knitr)
library(ggplot2)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

## 1.1

Name several entities, attributes, activities, events, and state variables for the following systems.

### (a) A cafeteria

#### Entities

* Serving Line
* Food Servers
* Tables

#### Attributes

* Number of Food Servers
* Number of seats per table
* Rate of serving for Food Servers
* Time range for eating the meal

#### Activities

* Waiting in line
* Being served by a Food Server
* Waiting for a table to eat
* Eating at a table

#### Events

* Arrival of new person in serving line to be served
* Person leaving serving line
* Person waiting for seat at table to eat
* Person finishing eating and leaving table

#### State Variables

* Number of people eating at tables
* Number of people waiting in line to be served

### (b) A grocery store

#### Entities

* Checkout lanes

#### Attributes

* Max number of items allowed in checkout lane
* Rate of checkout for cashier

#### Activities

* Customer shopping in the grocery store
* Customer checking out (paying for goods)

#### Events

* Arrival of customer at grocery store
* Arrival of customer at checkout lane
* Customer completing checkout
* Customer departing store without purchasing anything

#### State Variables

* Number of customers in grocery store
* Number of customers in checkout lane lines


### (c) A laundromat

#### Entities

* Washing machines
* Drying machines

#### Attributes

* Washing machine run time
* Drying machine run time
* Ratio of washing machine to drying machine capacity

#### Activities

* Washing clothes
* Drying clothes
* Loading washing machine
* Transfering from washing to drying machine
* Unloading from drying machine

#### Events

* Washing maching cycle starts
* Washing machine cycle stops
* Dryer cycle starts
* Dryer cycle stops

#### State Variables

* Number of busy washing machines
* Number of busy dryers

### (d) A fast-food restaurant

#### Entities

* Cashiers
* Back-cooks (i.e. burger flippers)
* Fryers

#### Attributes

* Burgers per burger flipper
* Orders of frys per Fryer
* Cashier busy or not

#### Activities

* Cooking a burger
* Making french frys
* Cashier taking order, accepting payment

#### Events

* Order in
* Order ready for pickup
* French fries done cooking

#### State Variables

* Number of orders pending
* Number of burgers being cooked
* Orders of french frys cooked/ready for serving
* Number of bugers being ready for serving

### (e) A hospital emergency room

#### Entities

* Doctors
* Beds
* Patients
* Admitting staff

#### Attributes

* Patients per Doctor


#### Activities

* Patient admitted
* Doctor take care of patient
* Patient discharged

#### Events

* Patient arrives
* Patient admitted
* Doctor discharges patient

#### State Variables

* Beds empty
* Patients awaiting addmission
* Paitents awaiting discharge


### (f) A taxicab company with 10 taxis

#### Entities

* Taxis
* Dispatcher
* Customers

#### Attributes

* Taxi has customer
* Taxi enroute to customer
* Customer waiting for taxi

#### Activities

* Enroute to customer
* Transporting customer

#### Events

* Picking up custommer
* Dropping off customer

#### State Variables

* Taxis with customers
* Customers waiting for available taxi

### (g) An automobile assembly line

#### Entities

* Parts
* Assembly machines
* Workers

#### Attributes

* Parts inventory
* Assembly machine rate of production
* Worker rate of production

#### Activities

* Machine assembling car 
* Worker assembling car
* Staging parts for use by Machine or Worker

#### Events

* Car assembly started
* Car assembly completed
* Parts depleted
* Car assembly by machine X completed
* Car assembly by worker Y completed

#### State Variables

* Cars on assembly line
* Parts inventory level
* Workers out sick/vacation
* Machines broken down

## 2.1 

Consider the following continuously operating job shop. Interarrival times of jobs are distributed as follows:

Time Between Arrivals (hours) | Probability
------------------------------|------------
0                             |  0.23
1                             |  0.37
2                             |  0.28
3                             |  0.12

Processing times for jobs are normally distributed, with mean 50 minutes, and standard deviation 8 minutes. Construct
a simulation table and perform a simulation for 10 new customers. Assume that, when the simulation starts, there is one
job being processed (scheduled to be completed in 25 minutes) and there is one job with a 50-minute processing time in
the queue.


```{r}
# Create a data frame of the pre-existing jobs
existingJobs <- data.frame(customer=c(-2, -1), 
                           iaHrs=c(0,0), 
                           iaMins=c(0,0), 
                           arrivalMins=c(0,0),
                           svcTimeMins=c(25, 50),
                           timeSvcBegin=c(0, 25),
                           queueWaitMins=c(0,25),
                           timeSvcEnd=c(25, 75),
                           timeInSystem=c(25,75))
# Create a data frame of the new customers and their jobs
newJobs <- data.frame(customer=seq(1, 10), 
                      iaHrs=c(0, sample(seq(0, 3), 
                                              size=9, 
                                              prob=c(.23, .37, .28, .12), 
                                              replace=TRUE)), 
                      iaMins=rep(NA, 10), 
                      arrivalMins=rep(0, 10),
                      svcTimeMins=rnorm(10, mean=50, sd=8),
                      timeSvcBegin=rep(0, 10),
                      queueWaitMins=rep(0, 10),
                      timeSvcEnd=rep(0, 10),
                      timeInSystem=rep(0, 10))
# Convert from interarrival hours to minutes and 
# determine overall arrival times
newJobs$iaMins <- newJobs$iaHrs * 60
newJobs$arrivalMins <- cumsum(newJobs$iaMins) 
# Join the existing and new jobs into one table
simTable <- rbind(existingJobs, newJobs)
# Loop over the rows the compute the various activity and clock times
for(i in seq(3, nrow(simTable)))
{
  simTable[i,]$timeSvcBegin <- max(simTable[i,]$arrivalMins, simTable[i-1,]$timeSvcEnd)
  simTable[i,]$queueWaitMins <- simTable[i,]$timeSvcBegin - simTable[i,]$arrivalMins
  simTable[i,]$timeSvcEnd <- simTable[i,]$timeSvcBegin + simTable[i,]$svcTimeMins
  simTable[i,]$timeInSystem <- simTable[i,]$timeSvcEnd - simTable[i,]$arrivalMins
}
# Show the table
kable(simTable)
```

#### (a) What was the average time in the queue for the 10 new jobs?

The average time in the queue for the 10 new jobs is computed below:

```{r}
mean(simTable[seq(3, 12),]$queueWaitMins)
```

#### (b) What was the average processing time of the 10 new jobs?

The average processing time is computed below:

```{r}
mean(newJobs$svcTimeMins)
```


#### (c) What was the maximum time in the system for the 10 new jobs?

The maximum time in the system for the 10 new jobs is computed below:

```{r}
max(simTable[seq(3, 12),]$timeInSystem)
```

## 2.2

A baker is trying to figure out how many dozens of bagels to bake each day. The probability distribution of the number of bagel customers is as follows:

Customer/Day | 8    | 10   | 12   | 14 
-------------|------|------|------|----
Probability  | 0.35 | 0.30 | 0.25 | 0.10

Customers order 1,2,3 or 4 dozen bagels according to the following probability distribution:

Dozen Ordered/Customer | 1   | 2   | 3   | 4   
-----------------------|-----|-----|-----|-----
Probability            | 0.4 | 0.3 | 0.2 | 0.1

Bagels sell for $8.40 per dozen. They cost $5.80 per dozen to make. All bagels not sold at the end of the day
are sold at half price to a local grocery store. Based on 5 days of simulation, how many dozen (to the nearest 5 dozen)
bagels should be baked each day?

```{r}
# Function to define a simulation at a specified 
# level of dozens of bagels produced.
bakersProfit <- function(bagelsMade)
{
  simDays <- 5
  revPerDoz <- 8.40
  costPerDoz <- 5.80
  simTable <- data.frame(day=seq(1, simDays),
                       customers=sample(c(8,10,12,14), 
                                        size=simDays, 
                                        prob=c(0.35, 0.30, 0.25, 0.10),
                                        replace=TRUE),
                       dozenOrdered=rep(NA, simDays),
                       revenue=rep(NA, simDays),
                       lostProfit=rep(NA, simDays),
                       salvage=rep(NA, simDays),
                       dailyCost=rep(NA, simDays),
                       dailyProfit=rep(NA, simDays))

  for(i in seq(1, nrow(simTable)))
  {
    bagelsOrdered <- sample(c(1,2,3,4),
                            size=simTable[i,]$customers,
                            prob=c(0.4, 0.3, 0.2, 0.1),
                            replace=TRUE)
    
    simTable[i,]$dozenOrdered <- sum(bagelsOrdered)
    
    simTable[i,]$revenue <- min(simTable[i,]$dozenOrdered, bagelsMade) * revPerDoz
    simTable[i,]$lostProfit <- max(simTable[i,]$dozenOrdered - bagelsMade, 0) * (revPerDoz - costPerDoz)
    simTable[i,]$salvage <- max(bagelsMade - simTable[i,]$dozenOrdered, 0) * (revPerDoz / 2)
    simTable[i,]$dailyCost <- bagelsMade * costPerDoz
    simTable[i,]$dailyProfit <- simTable[i,]$revenue + simTable[i,]$salvage - simTable[i,]$dailyCost
  }
  
  return(simTable)
}
# Loop over a range of dozens of bagels (0, 5, 10, etc)
dozens <- seq(0, 30, by=5)
profitTable <- data.frame(dozPerDay=c(), fiveDayProfit=c())
for(d in dozens)
{
  # Run the simulation for the given level of production
  simTable <- bakersProfit(d)
  profitTable <- rbind(profitTable, cbind(dozPerDay=d, fiveDayProfit=sum(simTable$dailyProfit)))
  #print(paste(d, " dozen/day: 5 day profit is ", profitTable[profitTable$dozPerDay==d,]$fiveDayProfit, ".", sep=""))  
}
```

The following table shows the profit associated with various levels of production:

```{r, echo=FALSE}
# Show the profit table at various dozens
kable(profitTable)
# Extract the row with max profit
profitMax <- profitTable[order(profitTable$fiveDayProfit, decreasing=TRUE)[1],]
```

The following table shows the details of the simulation for the maximum profit shown above (`r profitMax$dozPerDay` dozen bagels/day):

```{r, echo=FALSE}


# Re-run for the max profit dozens
maxProfitSimTable <- bakersProfit(profitMax$dozPerDay)
# Show the max profits sim table.
kable(maxProfitSimTable)
```


## 2.4

Smalltown Taxi operates one vehicle during the 9:00 A.M. to 5:00 P.M. period. Currently, consideration is being given to the addition of a second vehicle to the fleet. The demand for taxis follows the distribution shown:

Time Between Calls (minutes) | 15   | 20  | 25  | 30  | 35
-----------------------------|------|-----|-----|-----|-----
Probability                  | 0.14 | 0.22| 0.43| 0.17| 0.04

The distribution of time to complete a service is as follows:

Service Time (minutes) |   5   |  15  |  25  |  35  |  45 
-----------------------|-------|------|------|------|------
Probability            |  0.12 | 0.35 | 0.43 | 0.06 | 0.04

Simulate 5 individual days of operation of the current system and of the system with an 
additional taxicab. Compare the two systems with respect to the waiting times of the customers
and any other measures that might shed light on the situation.

### One Taxi Simulation

```{r}
# Function to wrap a single taxi cab on a single day
singleTaxiDailyCalls <- function(callsPerDay, maxDailyMinutes)
{
  # Create a data frame of the new customers and their jobs
  simTable <- data.frame(customer=seq(1, callsPerDay), 
                        iaMins=sample(seq(15, 35, by=5), 
                                                size=callsPerDay, 
                                                prob=c(0.14, 0.22, 0.43, 0.17, 0.04), 
                                                replace=TRUE), 
                        arrivalMins=rep(0, callsPerDay),
                        svcTimeMins=sample(seq(5, 45, by=10), 
                                                size=callsPerDay, 
                                                prob=c(0.12, 0.35, 0.43, 0.06, 0.04), 
                                                replace=TRUE),
                        timeSvcBegin=rep(0, callsPerDay),
                        queueWaitMins=rep(0, callsPerDay),
                        timeSvcEnd=rep(0, callsPerDay),
                        timeInSystem=rep(0, callsPerDay))
  
  # Determine overall arrival times
  simTable$arrivalMins <- cumsum(simTable$iaMins) 
  # Loop over the rows the compute the various activity and clock times
  for(i in seq(1, nrow(simTable)))
  {
    if(i == 1)
    {
      simTable[i,]$timeSvcBegin <- simTable[i,]$arrivalMins
         
    }
    else
    {
      simTable[i,]$timeSvcBegin <- max(simTable[i,]$arrivalMins, simTable[i-1,]$timeSvcEnd)
    }
    simTable[i,]$queueWaitMins <- simTable[i,]$timeSvcBegin - simTable[i,]$arrivalMins 
    simTable[i,]$timeSvcEnd <- simTable[i,]$timeSvcBegin + simTable[i,]$svcTimeMins
    simTable[i,]$timeInSystem <- simTable[i,]$timeSvcEnd - simTable[i,]$arrivalMins
  }  
  # Convert queue wait of zero to NA so we can aggregate only those who actually waited.
  simTable$queueWaitMins[simTable$queueWaitMins == 0] <- NA
  # subset to the max daily minutes in the business day
  simTable <- simTable[simTable$arrivalMins <= maxDailyMinutes,]
  
  return(simTable)
}
```

The following code segment executes the **single** taxi simulation over 5 days and aggregates the results. 

```{r}
# Run the single taxi simulation over 5 days
daysToSimulate <- 5
multiDaySimTable <- data.frame()
for(i in seq(1, daysToSimulate))
{
  oneDayOneTaxi <- singleTaxiDailyCalls(32, 480)
  multiDaySimTableOneTaxi <- rbind(multiDaySimTable, cbind(day=i, oneDayOneTaxi))
}
kable(multiDaySimTableOneTaxi)
# Show the table
kable(summary(multiDaySimTableOneTaxi[,c(2,7,9)]))
```

### 2 Taxi Simulation

Now lets develop the 2 taxi simulation.

```{r}
# Function to wrap a duel taxi cab company on a single day
duelTaxiDailyCalls <- function(callsPerDay, maxDailyMinutes)
{
  # Create a data frame of the new customers and their jobs
  simTable <- data.frame(customer=seq(1, callsPerDay), 
                        iaMins=sample(seq(15, 35, by=5), 
                                                size=callsPerDay, 
                                                prob=c(0.14, 0.22, 0.43, 0.17, 0.04), 
                                                replace=TRUE), 
                        arrivalMins=rep(0, callsPerDay),
                        tc1Avail=rep(0, callsPerDay),
                        tc2Avail=rep(0, callsPerDay),
                        taxiChosen=rep(0, callsPerDay),
                        svcTimeMins=sample(seq(5, 45, by=10), 
                                                size=callsPerDay, 
                                                prob=c(0.12, 0.35, 0.43, 0.06, 0.04), 
                                                replace=TRUE),
                        timeSvcBegin=rep(0, callsPerDay),
                        timeSvcEndTc1=rep(0, callsPerDay),
                        timeSvcEndTc2=rep(0, callsPerDay),
                        queueWaitMins=rep(0, callsPerDay),
                        timeInSystem=rep(0, callsPerDay))
  
  # Determine overall arrival times
  simTable$arrivalMins <- cumsum(simTable$iaMins) 
  # Loop over the rows the compute the various activity and clock times
  for(i in seq(1, nrow(simTable)))
  {
    if(i == 1)
    {
      simTable[i,]$timeSvcBegin <- simTable[i,]$arrivalMins
      simTable[i,]$taxiChosen <- 1
    }
    else
    {
      # Determine availability
      simTable[i,]$tc1Avail <- max(simTable[seq(1, i),]$timeSvcEndTc1)
      simTable[i,]$tc2Avail <- max(simTable[seq(1, i),]$timeSvcEndTc2)
      # Select taxi
      simTable[i,]$taxiChosen <- if (simTable[i,]$tc1Avail <= simTable[i,]$arrivalMins || 
                                     simTable[i,]$tc1Avail <= simTable[i,]$tc2Avail) 1 else 2      
      # Determine service start based on selected taxi
      simTable[i,]$timeSvcBegin <- if(simTable[i,]$taxiChosen == 1) max(simTable[i,]$arrivalMins, 
                                                                        simTable[i-1,]$tc1Avail) else max(simTable[i,]$arrivalMins, 
                                                                                                          simTable[i-1,]$tc2Avail)
    }
    simTable[i,]$timeSvcEndTc1 <- if(simTable[i,]$taxiChosen == 1) simTable[i,]$timeSvcBegin + simTable[i,]$svcTimeMins else 0
    simTable[i,]$timeSvcEndTc2 <- if(simTable[i,]$taxiChosen == 2) simTable[i,]$timeSvcBegin + simTable[i,]$svcTimeMins else 0
    simTable[i,]$queueWaitMins <- simTable[i,]$timeSvcBegin - simTable[i,]$arrivalMins 
    simTable[i,]$timeInSystem <- max(simTable[i,]$timeSvcEndTc1, simTable[i,]$timeSvcEndTc2) - simTable[i,]$arrivalMins
  }  
  # Convert queue wait of zero to NA so we can aggregate only those who actually waited.
  simTable$queueWaitMins[simTable$queueWaitMins == 0] <- NA
  # subset to the max daily minutes in the business day
  simTable <- simTable[simTable$arrivalMins <= maxDailyMinutes,]
  
  return(simTable)
}
```

The following table shows the execution of a single day for the **two** taxi simulation.

```{r, echo=FALSE}
kable(duelTaxiDailyCalls(32, 480))
```

The following code segment executes the two taxi simulation over 5 days and aggregates the results. 

```{r}
# Run the two taxi simulation over 5 days
daysToSimulate <- 5
multiDaySimTable <- data.frame()
for(i in seq(1, daysToSimulate))
{
  oneDayTwoTaxi <- duelTaxiDailyCalls(32, 480)
  multiDaySimTable <- rbind(multiDaySimTable, cbind(day=i, oneDayTwoTaxi))
}
#kable(multiDaySimTable)
# Show the table
kable(summary(multiDaySimTable[,c(2,12,13)]))
```

As can be seen in the column chart below, Taxi 1 has a much larger number of customers, 
but Taxi 2 does pick up approximately 20% of the customers.

```{r, echo=FALSE}
g1 <- ggplot(multiDaySimTable) + 
  geom_bar(aes(x=taxiChosen)) +
  labs(title="Distribution of Taxi Customers")  + myTheme
g1
```

## 2.5

The random variables X, Y, and Z are distributed as follows:

\[X ~ N (\mu = 100, \sigma^2=100) \]

\[Y ~ N (\mu = 300, \sigma^2=225) \]

\[Z ~ N (\mu = 40, \sigma^2=64) \]

Simulate 50 values of the random variable

\[W=\frac{X+Y}{Z}\]

Prepare a histogram of the resulting values, using class intervals of width equal to 3.

```{r}
fxX <- function (n)
{
  return (rnorm(n, 100, 10))
}

fxY <- function(n)
{
  return (rnorm(n, 300, 15))
}

fxZ <- function(n)
{
  return (rnorm(n, 40, 8))
}

n <- 50
dfVars <- data.frame(X=fxX(n), Y=fxY(n), Z= fxZ(n))
dfVars$W <- ( dfVars$X + dfVars$Y ) / dfVars$Z
```

The first 5 resulting random variables and the values of W are shown in the table below.

```{r, echo=FALSE}
kable(head(dfVars))
```

```{r, echo=FALSE}
g2 <- ggplot(dfVars) + geom_histogram(aes(x=W), binwidth=3) + myTheme
g2
```

## 2.7

Estimate, by simulation, the average number of lost sales per week for an inventory system that functions as follows:

(a) Whenever the inventory level falls to or below 10 unites, an order is placed. Only one order can be outstanding at a time.

(b) The size of each order is equal to 20 - I, where I is the inventory level when the order is placed.

(c) If a demand occurs during a period when the inventory level is zero, the sale is lost.

(d) Daily demand is normally distributed, with a mean of 5 units and a standard deviation of 1.5 unites. (Round off decimals to the closest integer during the simulation and, if a negative value results, give it a demand of zero.)

(e) Lead time is distributed  uniformly between zero and 5 days - integers only.

(f)  The simulation will start with 18 units in inventory.

(g) For simplicity, assume that orders are placed at the close of the business day and received after the lead time has occured. Thus, if lead time is one day, the order is available for distribution on the morning of the second day of business following the placement of the order.

(h)  Let the simulation run for 5 weeks.

```{r}
inventorySim <- function(daysToSimulate, daysPerWeek)
{
  simTable <- data.frame(day=seq(0, daysToSimulate),
                         #dayInCycle=c(daysPerWeek, rep(seq(1, daysPerWeek), daysToSimulate / daysPerWeek)),
                         beginInv=c(NA, rep(NA, daysToSimulate)),
                         demand=c(NA, round(rnorm(daysToSimulate, mean=5, sd=1.5))),
                         endInv=c(18, rep(NA, daysToSimulate)),
                         lostSales=rep(NA, daysToSimulate + 1),
                         #shortage=c(0, rep(NA, daysToSimulate)),
                         pendingOrder=c(0, rep(0, daysToSimulate)),
                         leadTime=c(NA, rep(NA, daysToSimulate)),
                         orderArriveDays=c(0, rep(0, daysToSimulate)))
  
  # Loop over the rows the compute the various activity and clock times
  for(i in seq(1, nrow(simTable)))
  {
    if(i == 1)
    {
      # Do nothing on first row (zeroth)
    }
    else
    {
      # Are any orders arriving today?
      pendingO <- 0
      if(simTable[i-1,]$orderArriveDays <= 1) 
      {
        pendingO <- simTable[i-1,]$pendingOrder
        simTable[i,]$pendingOrder <- 0
      }
      # Adjust begining inventory based on prior ending plus arriving orders
      simTable[i,]$beginInv <- simTable[i-1,]$endInv + pendingO
      if(simTable[i,]$demand > 0 && simTable[i,]$beginInv == 0)
      {
        simTable[i,]$lostSales <- simTable[i,]$demand
      }
      # Adjust ending inventory
      endI <- simTable[i,]$beginInv - simTable[i,]$demand #- simTable[i-1,]$shortage
      simTable[i,]$endInv <- max(endI, 0)
      
      # Ordering
      if(simTable[i,]$endInv <= 10 && simTable[i-1,]$orderArriveDays <= 1)
      {
        # New Order
        simTable[i,]$pendingOrder <- 20 - simTable[i,]$endInv
        simTable[i,]$leadTime <- round(runif(1, 0, 5))
        simTable[i,]$orderArriveDays <- simTable[i,]$leadTime
      }
      else
      {
        # Adjust arrival days for pending orders
        simTable[i,]$orderArriveDays <- if(simTable[i-1,]$orderArriveDays > 0) simTable[i-1,]$orderArriveDays - 1 else 0
        if(simTable[i,]$orderArriveDays > 0)
        {
          simTable[i,]$pendingOrder <- simTable[i-1,]$pendingOrder  
        }
      }
    }
  }
  
  return(simTable)
}

# Days based on 5 biz days/wk for n weeks
weeksToSim <- 5
daysToSim <- 5 * weeksToSim
inventorySimTable <- inventorySim(daysToSim)
```

```{r, echo=FALSE}
kable(inventorySimTable)
```

Based on the previously shown simulation data, the average lost sales per week 
are `r sum(inventorySimTable$lostSales, na.rm=TRUE) / weeksToSim`. 

```{r, echo=FALSE, warning=FALSE}
g1 <- ggplot(inventorySimTable) + 
  geom_point(aes(x=day, y=lostSales), colour="red") +
  geom_line(aes(x=day, y=beginInv), colour="lightblue") +
  labs(title="Lost Sales Simulation", y="Inventory Level") +
  myTheme 
g1
```

## 2.8

An elevator in a manufacturing plant carries exactly 400 kilograms of material. There are
three kinds of material packaged in boxes that arrive for a ride on the elavator. These materials
and their distributions of time between arrivals are as follows:

Material  | Weight (kg) | Interarrival Time  (minutes)
----------|-------------|-----------------------------
A         |  200        | 5 +- 2 (uniform)
B         |  100        | 6 (constant)
C         |   50        | P(2) = 0.33, P(3) = 0.67

It takes the elevator 1 minute to go up to the second floor, 2 minutes to unload, and 1 minute to return to the first floor.
The elevator does not leave the first floor unless it has a full load. Simulate 1 hour of operation of the system. What is the average transit time for a box of material A (time from its arrival until it is unloaded)? What ist he average waiting time for a box of material B? How many boxes of material C made the trip in 1 hour?

```{r}
# Function to wrap a single taxi cab on a single day
materialElevatorSim <- function(n, simMinutes, elevatorCapacity, travelUpMin, unloadMin, travelDownMin)
{
  # Material A interarrivals
  matAIA <- data.frame(mat=rep("A", n), 
                       weight=rep(200, n),
                       matIA=sample(c(3, 7), 
                        size=n, 
                        prob=c(0.50, 0.50), 
                        replace=TRUE))
  matAIA$arrivalTime <- cumsum(matAIA$matIA)   
  # Material B interarrivals                     
  matBIA <- data.frame(mat=rep("B", n), 
                       weight=rep(100, n),
                       matIA=rep(6, n))
  matBIA$arrivalTime <- cumsum(matBIA$matIA)   
  # Material C interarrivals
  matCIA <- data.frame(mat=rep("C", n), 
                       weight=rep(50, n),
                       matIA=sample(c(2, 3), 
                        size=n, 
                        prob=c(0.33, 0.67), 
                        replace=TRUE))
  matCIA$arrivalTime <- cumsum(matCIA$matIA)   
  
  materialArrivals <- rbind(matAIA, matBIA, matCIA)
  materialArrivals <- materialArrivals[with(materialArrivals, order(arrivalTime)),]
  
  simTable <- materialArrivals
  simTable$cumWeight <- rep(0, nrow(simTable))
  simTable$waiting <- rep(1, nrow(simTable))
  simTable$svcBeginTime <- rep(0, nrow(simTable))
  simTable$svcEndTime <- rep(0, nrow(simTable))
  simTable$waitTimeMins <- rep(0, nrow(simTable))
  simTable$transitTime <- rep(0, nrow(simTable))
  
  # Loop over the rows the compute the various activity and clock times
  for(i in seq(1, nrow(simTable)))
  {
    # Compute cumulative waiting weight
    if(i == 1)
    {
      simTable[i,]$cumWeight <- simTable[i,]$weight
    }
    else
    {
      cumWaitingWeight <- 0
      for(j in seq(1,i))
      {
        if(simTable[j,]$waiting == 1)
        {
          cumWaitingWeight <- cumWaitingWeight + simTable[j,]$weight
        }
      }
      simTable[i,]$cumWeight <- cumWaitingWeight 
    }
  
    # Determine riders this iteration
    if(simTable[i,]$cumWeight >= elevatorCapacity)
    {
      # Determine which items move on to elevator and which don't
      riders <- c()
      if(simTable[i,]$cumWeight == elevatorCapacity)
      {
        # Mark all waiters as riders
        for(j in seq(1, i))
        {
          if(simTable[j,]$waiting == 1)
          {
            riders <- c(riders, j)
          }
        }
      }
      else
      {
        # Find waiters that fit
        diff <- simTable[i,]$cumWeight - elevatorCapacity
        excluder <- 0
        for(j in seq(i, 1, by=-1))
        {
          if(simTable[j,]$waiting == 1 && simTable[j,]$weight == diff)
          {
            excluder <- j
            break;
          }
        }   
        
        for(j in seq(1, i))
        {
          if(simTable[j,]$waiting == 1 && j != excluder)
          {
            riders <- c(riders, j)
          }
        }            
      }
      
      for(k in riders)
      {
        simTable[k,]$waiting <- 0
        simTable[k,]$svcBeginTime <- max(simTable[i,]$arrivalTime, simTable[i,]$svcEndTime + travelDownMin)
        simTable[k,]$svcEndTime <- simTable[k,]$svcBeginTime + travelUpMin + unloadMin        
      }
    }
    
    #simTable[i,]$waitTimeMins <- 
    
    
#     simTable[i,]$queueWaitMins <- simTable[i,]$timeSvcBegin - simTable[i,]$arrivalMins 
#     simTable[i,]$timeSvcEnd <- simTable[i,]$timeSvcBegin + simTable[i,]$svcTimeMins
#     simTable[i,]$timeInSystem <- simTable[i,]$timeSvcEnd - simTable[i,]$arrivalMins
  }  
  # Convert queue wait of zero to NA so we can aggregate only those who actually waited.
  #simTable$queueWaitMins[simTable$queueWaitMins == 0] <- NA

  simTable$transitTime <- simTable$svcEndTime - simTable$arrivalTime
  simTable[simTable$transitTime < 0,]$transitTime <- NA
  
  simTable$waitTimeMins <- simTable$svcBeginTime - simTable$arrivalTime
  simTable[simTable$waitTimeMins < 0,]$waitTimeMins <- NA
  
  # Subset to the max simulation minutes 
  simTable <- simTable[simTable$arrivalTime <= simMinutes,]
  
  return(simTable)
}

materialSimTable <- materialElevatorSim(60, 60, 400, 1, 2, 1)
rownames(materialSimTable) <- seq(1, nrow(materialSimTable))
```


```{r, echo=FALSE}
kable(materialSimTable)
```

Average transit time for material A in computed below:

```{r}
transitTimeA <- mean(materialSimTable[materialSimTable$mat == "A",]$transitTime, na.rm=TRUE)
transitTimeA
```

Average waiting time for material B is computed below:

```{r}
waitTimeMinsB <- mean(materialSimTable[materialSimTable$mat == "B",]$waitTimeMins, na.rm=TRUE)
waitTimeMinsB
```

Number of  material C which received service is computed below:

```{r}
countCserviced <- nrow(materialSimTable[materialSimTable$mat == "C" & 
                                          materialSimTable$waiting == 0,])
countCserviced
```
