---
title: "Homework 4"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "March 22, 2016"
output: pdf_document
classoption: portrait
geometry: margin=0.5in
---
```{r, echo=FALSE, message=FALSE}
library(knitr)
#library(gplots)
library(ggplot2)
library(randtoolbox)
#library(gridExtra)
#library(plot3D)
#library(tseries)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

```{r, echo=FALSE, message=FALSE}
library(knitcitations)
library(RefManageR)

cleanbib()

cite_options(style="markdown")

bibPkgTseries <- bibentry(bibtype="Misc",
                 author=personList(person(family="Trapletti", given="Adrian"), 
                                   person(family="Hornik", given="Kurt")),
                 journal="Annual Review of Sociology",
                 title="tseries: Time Series Analysis and Computational Finance",
                 year=2015,
                 note="R package version 0.10-34",
                 url="http://CRAN.R-project.org/package=tseries")
```

## 1

*In this problem, you will implement and investigate a series of variance reduction procedures for Monte Carlo method by estimating the expected value of a cost function c(x) which depends on a D-dimensional random variable x.*

*The cost function is:*

\[c(x)=\frac{1}{(2\pi)^{\frac{D}{2}}} e^{-1/2 x^T x}\]

where

\[x_i \sim U(-5,5) \text{ for } i=1..D\]

Goal: estimate E[c(x)] - the expected value of c(x) - using Monte Carlo methods and see how it compares to the real value, which you are able to find by hand. 

```{r cost-function}
# First define the cost function as an R function
costFx <- function(x)
{
  b <- exp(-0.5 * t(x) * x)
  D <- length(x)
  res <- (1 / ((2 * pi)^(D/2))) * b
  return (res)
}
```

### a) Crude Monte Carlo

```{r crude-monte-carlo}
crudeMC <- function(n, min, max, d = 1)
{
  # Need a loop in here
  theta.hat <- matrix(nrow=d, ncol=n)
  for(i in 1:n)
  {
    x <- runif(d, min, max)
    theta.hat[,i] <- costFx(x)
    #gXbar <- (1/n) * costFx(x)
    #print(((max-min) * gXbar))
    #theta.hat[,i] <- t((max-min) * gXbar)
  }
  
  return (theta.hat)
}

#crudeMC(10, -5, 5, 2)

crudeMc.Loop <- function(d, verbose=FALSE)
{
  crudeMc.result <- data.frame(mean=c(), stdev=c(), n=c())
  for(n in seq(1000, 10000, by=1000))
  {
    res <- crudeMC(n=n, min=-5, max=5, d=d)
    if(verbose)
    {
      print("Data")
      print(res)
      print("Mean")
      print(mean(res))
    }
    
    crudeMc.result <- rbind(crudeMc.result, data.frame(mean=mean(res), stdev=sd(res), n=n))
  }
  
  crudeMc.result$EcActual <- (1/10)^d
  crudeMc.result$CoefVari <- crudeMc.result$stdev / crudeMc.result$mean
  
  return (crudeMc.result)
}
```

In the code below, we call the crude Monte Carlo loop function, show the top entries and visualize the result for D=1. The blue line represents the mean value, pink is the standard deviation, and the green line is the analytical value for \(E[c(x)] = (1/10)^D\).

```{r}
crudeMc.D1 <- crudeMc.Loop(d=1)
```

```{r, echo=FALSE, fig.height=3}
g1 <- ggplot(crudeMc.D1) + 
  geom_point(aes(x=n, y=mean), colour="lightblue") + 
  geom_line(aes(x=n, y=mean), colour="lightblue") + 
  geom_point(aes(x=n, y=stdev), colour="pink") + 
  geom_line(aes(x=n, y=stdev), colour="pink") + 
  geom_line(aes(x=n, y=EcActual), colour="lightgreen") + 
  labs(title="D=1 Monte Carlo Estimate by n", y="mean/stdev") +
  myTheme
g1
```

```{r, echo=FALSE}
kable(crudeMc.D1)
```


In the code below, we call the crude Monte Carlo loop function, show the top entries and visualize the result for D=2.

```{r}
crudeMc.D2 <- crudeMc.Loop(d=2, verbose=FALSE)
```

```{r, echo=FALSE, fig.height=3}
g1 <- ggplot(crudeMc.D2) + 
  geom_point(aes(x=n, y=mean), colour="lightblue") + 
  geom_line(aes(x=n, y=mean), colour="lightblue") + 
  geom_point(aes(x=n, y=stdev), colour="pink") +   
  geom_line(aes(x=n, y=stdev), colour="pink") +
  geom_line(aes(x=n, y=EcActual), colour="lightgreen") + 
  labs(title="D=2 Monte Carlo Estimate by n", y="mean/stdev") +
  myTheme
g1
```

```{r, echo=FALSE}
kable(crudeMc.D2)
```

### Quasi-Random Numbers

First, we compare the typical uniform random numbers from R's `runif` function to Sobol quasi-random numbers from `randtoolbox::sobol` function. 100 pairs of numbers are drawn from both generators and visualized below.

#### Uniform Random Numbers

The following code segment uses `runif` to generate \(m=100\) random numbers and plots them.

```{r}
m <- 100
unifRn <- as.data.frame(matrix(runif(m * 2), ncol=2))
colnames(unifRn) <- c("x", "y")
head(unifRn)
```

```{r, echo=FALSE, fig.height=3}
g2 <- ggplot(data=as.data.frame(unifRn)) + 
  geom_point(aes(x=x, y=y)) + 
  labs(title="Uniform Psuedo-Random Numbers n=100 x 2 (runif)") +
  myTheme
g2
```

#### Sobol Random Numbers

The following code segment uses `sobol` to generate \(m=100\) random numbers and plots them.


```{r}
sobolRn <- as.data.frame(sobol(m, d=2))
colnames(sobolRn) <- c("x", "y")
head(sobolRn)
```

```{r, echo=FALSE, fig.height=3}
g2 <- ggplot(data=as.data.frame(sobolRn)) + 
  geom_point(aes(x=x, y=y)) + 
  labs(title="Sobol Quasi-Random Numbers n=100 x 2 (randtoolbox::sobol)") +
  myTheme 
g2
```

At \(m=100\), the differences are less obvious, but there is some discernable pattern to the Sobol numbers which is more apparent at great \(m\). As such, the prefix "quasi" seems appropriate. The definition of "quasi" is "seemingly, apparently but not really". These might appear to be random numbers at first glace, but there is quite a pattern, the lattice, as more and more are generated.

#### Sobol Monte Carlo

First, `sobol` based helper functions are defined using the same structure as the prior `runif` based functions. A couple of changes were needed:

* `sobol` returns a matrix. This is converted to a vector for use in the cost function.
* `sobol` has an init parameter, but we don't want to re-initialize every call, so this is bubbled up to the loop function to allow it to drive the re-init. 

```{r sobol-monte-carlo}
sobolMC <- function(n, d = 1, init=TRUE)
{
  # Need a loop in here
  theta.hat <- matrix(nrow=d, ncol=n)
  for(i in 1:n)
  {
    x <- as.vector(sobol(n=1, d=d, init=init))
    theta.hat[,i] <- costFx(x)
    init <- FALSE # turn off the init for i > 1 iterations
  }
  
  return (theta.hat)
}

sobolMC(n=10, d=2)

sobolMc.Loop <- function(d, verbose=FALSE)
{
  initSobol <- TRUE
  sobolMC.result <- data.frame(mean=c(), stdev=c(), n=c())
  for(n in seq(1000, 10000, by=1000))
  {
    res <- sobolMC(n=n, d=d, initSobol)
    if(verbose)
    {
      print("Data")
      print(res)
      print("Mean")
      print(mean(res))
    }
    
    sobolMC.result <- rbind(sobolMC.result, data.frame(mean=mean(res), stdev=sd(res), n=n))
    initSobol <- FALSE # don't re-initialize sobol every time.
  }
  
  sobolMC.result$EcActual <- (1/10)^d
  sobolMC.result$CoefVari <- sobolMC.result$stdev / sobolMC.result$mean
  
  return (sobolMC.result)
}

```

In the code below, we call the Sobol Monte Carlo loop function, show the top entries and visualize the result for D=1. Again, the blue line represents the mean value, pink is the standard deviation, and the green line is the analytical value for \(E[c(x)] = (1/10)^D\).

```{r}
sobolMc.D1 <- sobolMc.Loop(d=1)
```

```{r, echo=FALSE, fig.height=3}
g1 <- ggplot(sobolMc.D1) + 
  geom_point(aes(x=n, y=mean), colour="lightblue") + 
  geom_line(aes(x=n, y=mean), colour="lightblue") + 
  geom_point(aes(x=n, y=stdev), colour="pink") + 
  geom_line(aes(x=n, y=stdev), colour="pink") + 
  geom_line(aes(x=n, y=EcActual), colour="lightgreen") + 
  labs(title="D=1 Sobol Monte Carlo Estimate by n", y="mean/stdev") +
  myTheme
g1
```

```{r, echo=FALSE}
kable(sobolMc.D1)
```



```{r}
sobolMc.D2 <- sobolMc.Loop(d=2) #data.frame(mean=c(), stdev=c(), EcActual=c(), n=c()) #
```

```{r, echo=FALSE, fig.height=3}
g1 <- ggplot(sobolMc.D2) + 
  geom_point(aes(x=n, y=mean), colour="lightblue") + 
  geom_line(aes(x=n, y=mean), colour="lightblue") + 
  geom_point(aes(x=n, y=stdev), colour="pink") + 
  geom_line(aes(x=n, y=stdev), colour="pink") + 
  geom_line(aes(x=n, y=EcActual), colour="lightgreen") + 
  labs(title="D=2 Sobol Monte Carlo Estimate by n", y="mean/stdev") +
  myTheme
g1
```

```{r, echo=FALSE}
kable(sobolMc.D2)
```
