---
title: "Homework 4"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "March 22, 2016"
output: pdf_document
classoption: portrait
geometry: margin=0.5in
---
```{r, echo=FALSE, message=FALSE}
library(knitr)
#library(gplots)
library(ggplot2)
#library(gridExtra)
#library(plot3D)
#library(tseries)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

```{r, echo=FALSE, message=FALSE}
library(knitcitations)
library(RefManageR)

cleanbib()

cite_options(style="markdown")

bibPkgTseries <- bibentry(bibtype="Misc",
                 author=personList(person(family="Trapletti", given="Adrian"), 
                                   person(family="Hornik", given="Kurt")),
                 journal="Annual Review of Sociology",
                 title="tseries: Time Series Analysis and Computational Finance",
                 year=2015,
                 note="R package version 0.10-34",
                 url="http://CRAN.R-project.org/package=tseries")
```

## 1

*In this problem, you will implement and investigate a series of variance reduction procedures for Monte Carlo method by estimating the expected value of a cost function c(x) which depends on a D-dimensional random variable x.*

*The cost function is:*

\[c(x)=\frac{1}{(2\pi)^{\frac{D}{2}}} e^{-1/2 x^T x}\]

where

\[x_i \sim U(-5,5) for i=1..D\]

Goal: estimate E[c(x)] - the expected value of c(x) - using Monte Carlo methods and see how it compares to the real value, which you are able to find by hand. 

```{r cost-function}
# First define the cost function as an R function
costFx <- function(x)
{
  b <- exp(-0.5 * t(x) * x)
  D <- length(x)
  res <- (1 / ((2 * pi)^(D/2))) * b
  return (res)
}
```

### a) Crude Monte Carlo

```{r crude-monte-carlo}
crudeMC <- function(n, min, max, d = 1)
{
  # Need a loop in here
  x <- runif(d, min, max)
  gXbar <- (1/n) * costFx(x)
  theta.hat <- (max-min) * gXbar
  
  return (theta.hat)
}


crudeMC(1000, -5, 5)

```