---
title: "Homework 2"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "February 21, 2016"
output: pdf_document
classoption: portrait
geometry: margin=0.5in
---
```{r, echo=FALSE}
library(knitr)
library(ggplot2)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

## 1

*Suppose that X is a discrete random variable having probability function \(Pr(X=k) = ck^2\) for \(k=1,2,3\). Find \(c\), \(Pr(X \leq 2)\), E[X] and Var(X).*

This suggests the following:

\[Pr(X=1) + Pr(X=2) + Pr(X=3) = 1\]
\[1^2c + 2^2c + 3^2c = 1\]
\[1c + 4c + 9c = 1\]
\[14c = 1\]
\[c = \frac{1}{14}\]

```{r}
# Define an R function for the probability function
prXk <- function(k)
{
  c = 1/14
  p <- c * k^2
  
  return (p)
}

prX1 <- prXk(1)
prX1
prX2 <- prXk(2)
prX2
prX3 <- prXk(3)
prX3

ExpVal <- prX1 + (2*prX2) + (3*prX3)

secondMoment <- prX1 + (2^2*prX2) + (3^2*prX3)
```

\(Pr(X \leq 2) = `r prX1 + prX2`\)

The expected value \(E[X] = 1 \times `r prX1` + 2 \times `r prX2` + 3 \times `r prX3` = `r ExpVal` \).

The variance \(Var(X)=E[X^2] - (E[X])^2= `r secondMoment` -  `r ExpVal^2` = `r secondMoment - ExpVal^2`\)

## 2

*Suppose that X is a continuous random variable having p.d.f \(f(x)=cx^2\) for \(1\leq x \leq 2\). Find \(c\), \(Pr(X \geq 1)\), E[X] and Var(X).*

\[\int cx^2 dx= c \int x^2 dx = c \times \frac{x^3}{3} + C \]

\[\int cx^2 dx= c \times \frac{x^3}{3} + C = 1\]

\[ c = \frac{3 \times 1}{x^3} = \frac{3}{x^3} \]




```{r}
pr2X <- function(x)
{
  c <- 3 / x^3
  return( c * x^2)
}

r <- seq(1, 2, by=0.1)

pr2Xr <- pr2X(r)

g1 <- ggplot(data=data.frame(x=r, y=pr2Xr)) + geom_point(aes(x=x, y=y))
g1

```