---
title: "Homework 2"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "February 21, 2016"
output: pdf_document
classoption: portrait
geometry: margin=0.5in
---
```{r, echo=FALSE}
library(knitr)
library(ggplot2)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

## 1

*Suppose that X is a discrete random variable having probability function \(Pr(X=k) = ck^2\) for \(k=1,2,3\). Find \(c\), \(Pr(X \leq 2)\), E[X] and Var(X).*

This suggests the following:

\[Pr(X=1) + Pr(X=2) + Pr(X=3) = 1\]
\[1^2c + 2^2c + 3^2c = 1\]
\[1c + 4c + 9c = 1\]
\[14c = 1\]
\[c = \frac{1}{14}\]

```{r}
# Define an R function for the probability function
prXk <- function(k)
{
  c = 1/14
  p <- c * k^2
  
  return (p)
}

prX1 <- prXk(1)
prX1
prX2 <- prXk(2)
prX2
prX3 <- prXk(3)
prX3

ExpVal <- prX1 + (2*prX2) + (3*prX3)

secondMoment <- prX1 + (2^2*prX2) + (3^2*prX3)
```

\(Pr(X \leq 2) = `r prX1 + prX2`\)

The expected value \(E[X] = 1 \times `r prX1` + 2 \times `r prX2` + 3 \times `r prX3` = `r ExpVal` \).

The variance \(Var(X)=E[X^2] - (E[X])^2= `r secondMoment` -  `r ExpVal^2` = `r secondMoment - ExpVal^2`\)

## 2

*Suppose that X is a continuous random variable having p.d.f \(f(x)=cx^2\) for \(1\leq x \leq 2\). Find \(c\), \(Pr(X \geq 1)\), E[X] and Var(X).*

\[\int cx^2 dx= c \int x^2 dx = c \times \frac{x^3}{3} + C \]

\[\int cx^2 dx= c \times \frac{x^3}{3} + C = 1\]

\[ c = \frac{3 \times 1}{x^3} = \frac{3}{x^3} \]


The probabilty \(Pr(X \geq 1)\) will equal 1 due to the fact the range is \(1\leq x \leq 2\).

Expected value is 

```{r}
pr2X <- function(x)
{
  c <- 3 / x^3
  return( c * x^2)
}

r <- seq(1, 2, by=0.01)

pr2Xr <- pr2X(r)
```

```{r, fig.height=2.5}
g1 <- ggplot(data=data.frame(x=r, y=pr2Xr)) + geom_point(aes(x=x, y=y))
g1
```


## 3

Suppose the X and Y are jointly continuous random variables with 

\[y-x for 0 < x < 1 \text{ and } 1 < y < 2 \]
\[0 \text{ otherwise} \]

## 4

*Suppose that the following 10 obserations come from some distribution (not highly skewed) with unknown mean \(\mu\).*

```{r}
d <- c(7.3,6.1,3.8,8.4,6.9,7.1,5.3,8.2,4.9,5.8)
kable(t(d))
```

*Compute \(\bar{X}, S^2\) and an approximate 95% confidence interval for \(\mu\).*

```{r}
Xbar <- mean(d)
varD <- var(d)
sdD <- sd(d)
n <- length(d)
seD <- sdD / sqrt(n)
tval <- qt(.975, df=n-1)
```

\(\bar{X}=`r Xbar`\)

\(S^2=`r varD`\)

\(SE=`r seD`\)

\(\text{t-value } = `r tval`\)

95% CI of \(\mu\) = \(\bar{X} \pm t \times SE\) = (`r Xbar - (tval * seD)`, `r Xbar + (tval * seD)`)

# 5

*A random variable X has the memoryless property if, for all s, t > 0,*

\[Pr(X > t + s |X > t) = Pr(X > s)\]

*Show that the exponential distribution has the memoryless property.*

Page 193 of the DES text has an excellent proof of the the memoryless property:

\[P(X > t + s |X > t) = \frac{P(X > t + s)}{P(X > t)}\]

\[P(X > t + s |X > t) = \frac{e^{-\lambda(t+s)}}{e^{-\lambda t}}=e^{-\lambda s} = P(X > s)\]

# 6

Suppose \(X_1, X_2,...,X_n\) are i.i.d. Exp(\(\lambda=1\)). Use the Central Limit Theorem to find the approximate value of \(Pr(100 \leq \Sigma^{100}_{i=1} X_i \leq 110\)

We can take many n=100 samples from the Exp(\(lambda=1\)) distribution, and then sum them to create a distribution of the sums. Finally, use the number of sum samples that fall in the designated range divided by the total number of the n=100 samples to find an approximate probability.

```{r}
sumX <- c()
for(j in 1:10000)
{
 X <- rexp(100)
 sX <- sum(X)
 sumX <- c(sumX, sX)
}
```


```{r, fig.height=3}
g1 <- ggplot(data.frame(sumX=sumX)) + geom_histogram(aes(x=sumX)) + myTheme
g1
```

```{r}
n <- length(sumX)
valInRange <- length(sumX[100 <= sumX & sumX <= 110])

prInRange <- valInRange / n
```

The approximate value of \(Pr(100 \leq \Sigma^{100}_{i=1} X_i \leq 110\)) = `r prInRange`