---
title: "Homework 2"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "February 21, 2016"
output: pdf_document
classoption: portrait
geometry: margin=0.5in
---
```{r, echo=FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
library(plot3D)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

## 1

*Suppose that X is a discrete random variable having probability function \(Pr(X=k) = ck^2\) for \(k=1,2,3\). Find \(c\), \(Pr(X \leq 2)\), E[X] and Var(X).*

This suggests the following:

\[Pr(X=1) + Pr(X=2) + Pr(X=3) = 1\]
\[1^2c + 2^2c + 3^2c = 1\]
\[1c + 4c + 9c = 1\]
\[14c = 1\]
\[c = \frac{1}{14}\]

```{r}
# Define an R function for the probability function
prXk <- function(k)
{
  c = 1/14
  p <- c * k^2
  
  return (p)
}

prX1 <- prXk(1)
prX1
prX2 <- prXk(2)
prX2
prX3 <- prXk(3)
prX3

ExpVal <- prX1 + (2*prX2) + (3*prX3)

secondMoment <- prX1 + (2^2*prX2) + (3^2*prX3)
```

\(Pr(X \leq 2) = `r prX1 + prX2`\)

The expected value \(E[X] = 1 \times `r prX1` + 2 \times `r prX2` + 3 \times `r prX3` = `r ExpVal` \).

The variance \(Var(X)=E[X^2] - (E[X])^2= `r secondMoment` -  `r ExpVal^2` = `r secondMoment - ExpVal^2`\)

## 2

*Suppose that X is a continuous random variable having p.d.f \(f(x)=cx^2\) for \(1\leq x \leq 2\). Find \(c\), \(Pr(X \geq 1)\), E[X] and Var(X).*

\[\int cx^2 dx= c \int x^2 dx = c \times \frac{x^3}{3} + C \]

\[\int cx^2 dx= c \times \frac{x^3}{3} + C = 1\]

\[ c = \frac{3 \times 1}{x^3} = \frac{3}{x^3} \]


The probabilty \(Pr(X \geq 1)\) will equal 1 due to the fact the range is \(1\leq x \leq 2\).

Expected value is... hmmm...

The following visualization shows the probability density function:

```{r}
# Define an R funciton for the PDF
pr2X <- function(x)
{
  c <- 3 / x^3
  return( c * x^2)
}

r <- seq(1, 2, by=0.01)

pr2Xr <- pr2X(r)
```

```{r, fig.height=2.5}
g1 <- ggplot(data=data.frame(x=r, y=pr2Xr)) + geom_point(aes(x=x, y=y))
g1
```


## 3

*Suppose the X and Y are jointly continuous random variables with*

\[y-x \text{ for } 0 < x < 1 \text{ and } 1 < y < 2 \]
\[0 \text{ otherwise} \]

Assuming \(f_{X,Y}(x,y)=y-x\) is the joint density according to the above constraints:

### a) Compute and plot \(f_X(x)\) and \(f_Y(y)\)

\(f_X(x)\) and \(f_Y(y)\) are the marginal densities, therefore

\(f_X(x) = \int f_{X,Y}(x,y)dy=\frac{1}{2}y^2 - xy\) 

\(f_Y(y) = \int f_{X,Y}(x,y)dx=xy -\frac{1}{2}x^2\) 

```{r}
f_X <- function(x,y)
{
  return ((1/2 * y^2) - (x*y))
}
f_Y <- function(x,y)
{
  return ((x*y) - (1/2 * x^2) )
}

x <- seq(0, 1, by=0.01)
y <- seq(1, 2, by=0.01)
M <- mesh(x, y)
surf3D(M$x, M$y, z=f_X(M$x, M$y),  bty = "f", main="f_X Visualization")


x <- seq(0, 1, by=0.01)
y <- seq(1, 2, by=0.01)
M <- mesh(x, y)
surf3D(M$x, M$y, z=f_Y(M$x, M$y),  bty = "f", main="f_Y Visualization")

```

### b) Are X and Y independent?

if \(f_{X,Y}(x,y)=f_X(x)f_Y(y)\), we say X, Y are independent.

\(f_X(x)f_Y(y) = (\frac{1}{2}y^2 - xy)(xy -\frac{1}{2}x^2) = \frac{1}{2}xy^3-\frac{5}{4}x^2y^2+\frac{1}{2}x^3y\)

The above is not equal to \(y-x\), therefore X and Y are not independent.

### c) Compute \(F_X(x)\) and \(F_Y(y)\)

\(F_X(x)=\int f_X(x)dx = \int \frac{1}{2}y^2 - xy dx = \int \frac{1}{2}y^2 dx - \int xy dx = \frac{1}{2}xy^2 - \frac{x^2y}{2}\)


\(F_Y(y)=\int f_Y(y)dy = \int xy -\frac{1}{2}x^2 dy = \int xy dy - \int \frac{1}{2}x^2 dy = \frac{xy^2}{2} - \frac{1}{2}yx^2 \)


### d) Compute E[X], Var(X), E[Y], Var(Y), Cov(X,Y), and Corr(X,Y)

...



## 4

*Suppose that the following 10 obserations come from some distribution (not highly skewed) with unknown mean \(\mu\).*

```{r}
d <- c(7.3,6.1,3.8,8.4,6.9,7.1,5.3,8.2,4.9,5.8)
kable(t(d))
```

*Compute \(\bar{X}, S^2\) and an approximate 95% confidence interval for \(\mu\).*

```{r}
Xbar <- mean(d)
varD <- var(d)
sdD <- sd(d)
n <- length(d)
seD <- sdD / sqrt(n)
tval <- qt(.975, df=n-1)
```

\(\bar{X}=`r Xbar`\)

\(S^2=`r varD`\)

\(SE=`r seD`\)

\(\text{t-value } = `r tval`\)

95% CI of \(\mu\) = \(\bar{X} \pm t \times SE\) = (`r Xbar - (tval * seD)`, `r Xbar + (tval * seD)`)

# 5

*A random variable X has the memoryless property if, for all s, t > 0,*

\[Pr(X > t + s |X > t) = Pr(X > s)\]

*Show that the exponential distribution has the memoryless property.*

Page 193 of the DES text has an excellent proof of the the memoryless property:

\[P(X > t + s |X > t) = \frac{P(X > t + s)}{P(X > t)}\]

\[P(X > t + s |X > t) = \frac{e^{-\lambda(t+s)}}{e^{-\lambda t}}=e^{-\lambda s} = P(X > s)\]

# 6

Suppose \(X_1, X_2,...,X_n\) are i.i.d. Exp(\(\lambda=1\)). Use the Central Limit Theorem to find the approximate value of \(Pr(100 \leq \Sigma^{100}_{i=1} X_i \leq 110\)

We can take many n=100 samples from the Exp(\(lambda=1\)) distribution, and then sum them to create a distribution of the sums. Finally, use the number of sum samples that fall in the designated range divided by the total number of the n=100 samples to find an approximate probability.

```{r}
sumX <- c()
for(j in 1:10000)
{
 X <- rexp(100)
 sX <- sum(X)
 sumX <- c(sumX, sX)
}
```


```{r, fig.height=3, warning=FALSE}
g1 <- ggplot(data.frame(sumX=sumX)) + 
  geom_histogram(aes(x=sumX)) +
  labs(title="Distribution of Sums of n=100 Samples from Exp(lambda=1)") +
  myTheme
g1
```

```{r}
n <- length(sumX)
valInRange <- length(sumX[100 <= sumX & sumX <= 110])

prInRange <- valInRange / n
```

The approximate value of \(Pr(100 \leq \Sigma^{100}_{i=1} X_i \leq 110\)) = `r prInRange`


# 5.13

*A random variable X that has pmf given by \(p(x)=1/(n+1)\) over the range \(R_X=(0,1,2,...,n)\) is said to have a discrete uniform distribution.*

*(a) Find the mean and variance of this distribution.*

The mean would be \(\Sigma_{all i} i p(x)\). In this case \(p(x)\) is uniform at \(\frac{1}{n+1}\). As a result, this can be factored out of the mean equation resulting in \(p(x) \Sigma_{all i} i \). The hint told us that \(\Sigma_{all i} i =\frac{n(n+1)}{2}\), therefore the mean would be:

\[\text{mean }\ = \frac{1}{n+1} \times \frac{n(n+1)}{2} = \frac{n}{2}\]

The variance works the same way for discrete distributions, therefore:

\[\text{variance } = (\frac{1}{n+1} \times \frac{n(n+1)(2n+1)}{6}) - (\frac{n}{2})^2= \frac{n(2n+1)}{6} - \frac{n^2}{4} \]

\[= \frac{2n(2n+1)}{12} - \frac{3n^2}{12} = \frac{2n(2n+1)-3n^2}{12} = \frac{n^2 + 2n}{12}=\frac{n(n+2)}{12}\]

```{r, echo=FALSE, eval=FALSE}
# \[\text{variance } = \frac{1}{n+1} \times \frac{n(n+1)(2n+1)}{6} = \frac{n(2n+1)}{6}\]

# Empirical test, not shown or executed
n <- 4
pX <- function(n) {1/(n+1)}
pX

df <- data.frame(n=c(), mu=c(), half=c())
for(n in 1:20)
{
  mu <- 0
  for(i in 1:n)
  {
    mu <- mu + (i * pX(n))
  }
  
  df <- rbind(df, cbind(n=n, mu=mu, half=n/2))
}

df$match <- round(df$mu,3) == round(df$half,3)
df
```

*(b) If \(R_X = (a, a+1, a +2, ..., b)\), compute the mean and variance of X.*

Assuming n=b-a and therefore p(x)=1/(n+1) would still apply as in \(R_X=(2,3,4,5), a=2, b=5, n=3 \text{ and } p(x)=1/(3+1)=1/4\) to retain the uniform distribution. Also refering to http://mathforum.org/library/drmath/view/57166.html, the sum of \(R_X\) would be \(k(2a+k-1)/2\) with k=# of items=n+1.

\[\text{mean }\ = \frac{1}{n+1} \times \frac{n(2a+n-1)}{2} = \frac{n(2a+n-1)}{2(n+1)}\]

Variance, hmm....



```{r, echo=FALSE, eval=FALSE}
s <- 3
n <- 4

v <- 0
for(i in seq(s, s+n))
{
    v <- v + ( i^2 * (1/n+1))
}

v
```

# 5.14

*The lifetime in years of a satellite placed in orbit is given by the following pdf:*

\[f(x) = 0.4e^{-0.4x}, x \geq 0\]


*(a) What is the probability that this satellite is still "alive" after 5 years?*

```{r, fig.height=3}
# Define CDF and PDF functions
cdf514 <- function(x)
{

  p <- 1 - exp(-0.4 * x)

  return (p)
}

pdf514 <- function(x)
{

  p <- 0.4 * exp(-0.4 * x)

  return (p)
}

# prepare data set to help us visualize
rx <- data.frame(x=seq(0, 10, by=0.5))
rx$cdf <- cdf514(rx$x)
rx$pdf <- pdf514(rx$x)
# Visualize
g1 <- ggplot(rx) + geom_line(aes(x=x, y=pdf)) + myTheme
g2 <- ggplot(rx) + geom_line(aes(x=x, y=cdf)) + myTheme
grid.arrange(g1, g2, ncol=2)

p5 <- pdf514(5)
```

The probability that this satellite is still alive after 5 years is `r p5`.

*(b) What is the probability that the satellite dies between 3 and 6 years from the time it is placed in orbit?*

```{r}
# Compute probabilty of satelite's life time is between 3-6 years (i.e. it dies in this range)
p36 <- cdf514(6) - cdf514(3) 
```

The probabilty that this satellite dies between 3 and 6 years is `r p36`.


# 5.39

...


# Useful Links

Jointly continuous random variables: http://math.arizona.edu/~tgk/464_10/chap6_10_18.pdf