---
title: "Homework 3"
subtitle: "DATA604 Simulation and Modeling"
author: "Daniel Dittenhafer"
date: "March 6, 2016"
output: pdf_document
classoption: portrait
geometry: margin=0.5in
---
```{r, echo=FALSE}
library(knitr)
library(gplots)
library(ggplot2)
library(gridExtra)
library(plot3D)
library(tseries)
set.seed(20275)
# My ggplot theme
myTheme <- theme(axis.ticks=element_blank(),
                 axis.title=element_text(size="10"),
                  panel.border = element_rect(color="gray", fill=NA), 
                  panel.background=element_rect(fill="#FBFBFB"), 
                  panel.grid.major.y=element_line(color="white", size=0.5), 
                  panel.grid.major.x=element_line(color="white", size=0.5),
                  plot.title=element_text(size="10"))
```

```{r}
library(knitcitations)
library(RefManageR)

cleanbib()

cite_options(style="markdown")

bibPkgTseries <- bibentry(bibtype="Misc",
                 author=personList(person(family="Trapletti", given="Adrian"), 
                                   person(family="Hornik", given="Kurt")),
                 journal="Annual Review of Sociology",
                 title="tseries: Time Series Analysis and Computational Finance",
                 year=2015,
                 note="R package version 0.10-34",
                 url="http://CRAN.R-project.org/package=tseries")
```

## 1

*Starting with \(X_0=1\), write down the entire cycle for \(X_i=11X_{i-1}\text{ mod}(16)\)*

```{r}
fn1 <- function(x0)
{
  df <- data.frame(X=c(), R=c())  
  x <- x0
  continue <- TRUE
  
  while(continue)
  {
    xi <- (11 * x) %% 16
    df <- rbind(df, data.frame(X=x, R=xi))
    x <- xi
    
    if(xi == x0)
    {
      break
    }
  }
  
  return(df)
}

res <- fn1(1)
```

```{r, echo=FALSE}
kable(res)
```


## 2

*Using the LCG provided below: \(X_i=(X_{i-1} + 12) mod(13)\), plot the pairs \((U_1.U_2), (U_2, U_3),...\) and observe the lattice structure obtained. Discuss what you observe.

```{r}
fn2 <- function(x0, max=100)
{
  df <- data.frame(U1=c(), U2=c())  
  x <- x0
  
  for(i in 1:max)
  {
    xi <- (x + 12) %% 13
    df <- rbind(df, data.frame(U1=x, U2=xi))
    x <- xi
  }
  
  return(df)  
}
# Call the function starting at x0=1
res <- fn2(1)
```


```{r, echo=FALSE, fig.height=3, fig.width=3}
g21 <- ggplot(res) + geom_point(aes(x=U1, y=U2), alpha=0.25) + myTheme
g21
```

The chart above suggests there are only 13 points, but actually the LCG cycle period is 13 and numbers are repeating.

```{r, echo=FALSE}
kable(head(res, 14))
```

## 3

*Implement the pseudo-random number generator:*

\[X_i = 16807X_{i-1} \text{ mod}(2^{31} - 1)\]

*Using the seed 1234567, run the generator for 100,000 observations. Perform a chi-square goodness-of-fit test on the resulting PRN's. Use 20 equal-probability intervals and level \(\alpha = 0.05\). Now perform a runs up-and-down test with \(\alpha = 0.05\) on the observvations to see if they are independent.*

```{r}
fnLCG3 <- function(seed = 1, n = 1)
{
  rands <- rep(NA, n) 
  x <- seed
  modVal <- (2^31 - 1)
  
  for(i in 1:n)
  {
    xi <- (16807 * x) %% (modVal)
    rands[i] <- xi
    x <- xi
  }
  
  return(rands)  
}

n=100000
rn <- fnLCG3(1234567, n)
```

```{r, echo=FALSE}
kable(head(rn))
```

```{r, echo=FALSE, fig.height=3, message=FALSE}
#g3 <- ggplot(data.frame(random=rn)) + geom_histogram(aes(x=random), bins=20) + myTheme
#g3
```

### Chi-Square Test

```{r}
intervals <- 20
maxRn <- max(rn)
minRn <- min(rn)
intWidth <- (maxRn - minRn) / intervals
lwr <- minRn
dfCounts <- data.frame(intID=c(), count=c())
# Bin the data ourselves, I'd guess there 
# is an easier way, but this will do.
for(i in 1:intervals)
{
  upr <- lwr + intWidth
  inRange <- rn[lwr <= rn & rn < upr]
  dfCounts <- rbind(dfCounts, data.frame(intID=i, count=length(inRange)))
  # setup for next interval range
  lwr <- upr
}
# Do our own Chi-Squared test
Expected <- (100000 / intervals)
chi2 <- sum((dfCounts$count - Expected)^2 / Expected)
chi2
# Use built-in function to compare
chiTest <- chisq.test(dfCounts$count)
chiTest
```

The p-value = `r chiTest$p.value` is not less than \(\alpha = 0.05\), therefore we doesn't reject the null hypothesis that the distrubtion is uniform. 

```{r, echo=FALSE}
kable(dfCounts)
```

### Runs Up-and-Down Test

Using the `tseries` package, we execute the Runs test `r citep(bibPkgTseries)`. First we have to construct the +/- vector. Here we simply convert to boolean.

```{r}
s <- rep(NA, n - 1)
for(i in 1:n - 1)
{
  s[i] <- rn[i] < rn[i + 1]
}

runsTest <- runs.test(as.factor(s))
runsTest
```

Based on the p-value < 0.05, we reject the null hypothesis and conclude there is not evidence to support independence in the psuedo-random numbers.


## 4. 

*Give inverse-transforms, composition, and acceptance-rejection algorithms for generating from the following density:*

\[f(x)=
  \begin{cases} 
    \frac{3x^2}{2} & -1 \leq x \leq 1 \\
    0              & otherwise  \\
  \end{cases}
\]

### Inverse-Transforms

First find the definite integral of the probability density function:

\[F(x)=\int\frac{3x^2}{2} dx = \frac{x^3}{2}\]

Next set \(F(x)=R\) and solve for x in terms of R:

\[\frac{x^3}{2} = R\]

\[x^3 = 2R\]

\[x = \sqrt[3]{2R}\]

```{r}
# Define a function of the F^-1(X)
invTfn4 <- function(r)
{
  vals <- (2 * r)^(1/3)
  return (vals)
}

# Generate the uniform psuedo-random vars
rVals <- runif(n, -1, 1)
# Convert to the desired distribution using the inverse transform method.
invTVals <- invTfn4(rVals)
```

```{r, echo=FALSE, fig.height=3}
g41 <- ggplot(data.frame(invT=invTVals)) + geom_histogram(aes(x=invT)) + 
  myTheme +
  xlim(-1, 1) +
  labs(title="Empircial Distribution from Inverse Transform Method")

g41
```

### Acceptance-Rejection Method

```{r}
# Helper function for the Accept/Reject approach
myRARmethod <- function(fun, min, max)
{
  M <- 2
  accepted <- FALSE
  while(!accepted)
  {
    # Get a random value from uniform distrubtion (g(x) for us)
    r <- runif(1, min, max)

    # Sample x from g(x) and u from U(0,1) (the uniform distribution over the unit interval)
    u <- runif(1, 0, 1)
    gx <- dunif(r, min, max)

    # Check whether or not u<f(x)/Mg(x). 
    if(u < fun(r) / (M * gx))
    {
      accepted = TRUE
    }
  }

  return(r)
}

# Define a function for the PDF
Arfn4 <- function(x)
{
  if(-1 <=x && x <= 1)
  {
    val <- (3 * x^2) / 2
  }
  else
  {
    val = 0
  }
  return (val)
}

# Loop to generate the values
rarVals <- rep(NA, n)
for(i in 1:n)
{
  rarVals[i] <- myRARmethod(Arfn4, -1, 1)  
}

```

```{r, echo=FALSE, fig.height=3}

g42 <- ggplot(data.frame(ar=rarVals)) + geom_histogram(aes(x=ar)) + myTheme +
  labs(title="Accept-Reject Methods")
g42
```


## References


```{r, results='asis', echo=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```
  